### 2019-06-07 2-Chapters/2-Chapter/chapter2.tex
Typo

Multi-Armed Bandits (MAB) models were introduced by Thompson as early as in 1933 \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52} and others.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
Their name refers to one-armed bandits found in casinos, as shown in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX

% In all this thesis, we only consider one-dimensional exponential families, and in particular Bernoulli distributions.
We formally define the model and the main notations used in all this document in Section~\ref{sec:2:notations}.
%
Multi-armed bandits have been applied to various fields, from recommender systems and information retrieval to healthcare and finance. We discuss some applications in Section~\ref{sec:2:applicationsofStochasticMAB}.
In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits has already been applied with success in the last ten years.

@@

A rich research literature has produced many different algorithms tackling the MAB problem, as suggested by the length of the reference book \cite{LattimoreBanditAlgorithmsBook}.
We give in Section~\ref{sec:2:famousMABalgorithms} an overview of the main families of algorithms designed to tackle stochastic MAB problems, and we mainly focus on the $\UCB_1$, \klUCB{} and Thompson sampling algorithms, as they are used in the rest of the manuscript.
% We present SMPyBandits in the next Chapter~\ref{chapter:3}, a Python open-source library, designed to run numerical simulations of MAB problems, one of the main contributions of this thesis.
% We use it to compare the empirical performances of the most famous and most efficient algorithms, in terms of regret and by evaluating them on different problems.
% %
% In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
% We also empirically compare different algorithms for these two measures in Section~\ref{sec:3:timeAndMemoryCosts}.
%
% We then give regret upper-bounds of the two main algorithms used in this thesis, $\UCB_1$ and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They are order-optimal as they attain a logarithmic regret.
% proving their order optimality.
% (\ie, their regret upper-bound asymptotically matches the lower-bound up-to a constant factor).
% Moreover, \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.

Finally, we explore in Section~\ref{sec:2:chooseYourPreferredBanditAlgorithm} the problem of algorithm selection: as there are many different algorithms, all having different qualities and properties, how one can decide which algorithm to use for a particular application?
We present one of the first contributions made during my PhD \cite{Besson2018WCNC}, consisting of an efficient algorithm for aggregating algorithms,
and we illustrate its performance empirically.
% , but no satisfactory theoretical result is given, as the bound we had hoped for has been proved unachievable since then.

@@

A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the selected action, also called a \emph{reward}.
\textbf{We always consider $K\in\N, K\geq2$ arms, and discrete times $t\in\N^*$}.
When the player decides at time $t$ to play (or pull) the arm $A(t)\in\{1,\dots,K\}=[K]$, she receives a reward $r(t) = Y_{A(t),t}\in\R$, and so on for a finite number of steps $t$, from $t=1$ until $t=T$ for an \emph{horizon} $T$.
%
A commonly studied goal for the player is to maximize its sum of received rewards, $\sum_{t=1}^T r(t)$
(or its \emph{mean} in stochastic and stationary models).

@@

The most common objective for the player is to maximize its expected rewards $\E[\sum_{t=1}^T r(t)] = \sum_{t=1}^T \E[Y_{A(t),t}]$.
The expectancy is taken on the rewards $(Y_{k,t})_{k,t}$, and the extern random variables $(U_s)_s$ (\ie, on the algorithm's decisions).
%
Other commonly studied objectives include best-arm identification (BAI) \cite{audibert2010best}.

@@

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $100$ steps, $\sum_{t=1}^{100} r(t)$.
Your decisions have to be made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r(t)\in\{0,1\}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated Probs'' line.

In the first Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem}, the current state of the game is shown at time $t=24$.
% Currently at time $t=24$ (out of $T=100$ total time steps),
At this step, the player has collected a sum of rewards of $14$, by observing $X(t) = [6,2,2,2,2]$ rewards of value $1$ in the $K=5$ different arms. Arms were sampled $N(t) = [8,4,4,4,4]$ times, meaning that the value $0$ was seen respectively $[2,2,2,2,2]$ times, and currently arm $1$ appears to be the best one. The true means of the arms are $\bm{\mu}=[0.6, 0.2, 0.55, 0.7, 0.5]$, and (much) more samples are needed before the player can accurately identify arm $4$ as the best arm.
%
In the second Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem__step100} below, we display the result of an example of run, when the player was following the $\UCB_1$ algorithm from \cite{Auer02} (see Section~\ref{sub:2:IndexPolicies}).
After $T=100$ steps, the player obtained a cumulated reward of $56$, by playing mostly arms $4,3,5,1,0$ (in decreasing order of number of plays). The empirical means $\widehat{\mu_k}(T)$ correctly identify the best arm (arm $4$), but do not correctly rank the arms as arms $1$ and $3$ obtained means of $\widehat{\mu_1}(T) = 0.5 < \widehat{\mu_3}(T)=0.6$ but the true means are $\mu_3 = 0.55 < \mu_1 = 0.6$.
Other examples of such results, for different algorithms, and $T=100$ or $T=10000$, are given in Section~\ref{sub:2:shortNumericalExperiments}.

@@

A \emph{bandit algorithm} $\cA$ is also referred to as a strategy or a policy, and sometimes it is denoted by $\pi$ or $\rho$ in the literature.
The algorithm $\cA$ selects an arm $A(t)$ at time $t$, possibly by using the past observations and the past external randomness.
As shown below, being oblivious to the past yields very poor performance (cf. the pure exploration policy in Section~\ref{sub:2:naiveSimpleStrategies}), so efficient policies indeed depend on the successive feedbacks.
%
More formally, an algorithm can be defined as a sequence of \emph{measurable} functions $(\cA_t)_{t\geq1}$,
where $\cA_t$ maps the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$
to an arm $\cA_t(O_t) = A(t) \in[K]$
(we remind that we denote $r(s) = Y_{A(s),s}$ the $s$-th reward).
The initial information is reduced to $O_1 = (U_0)$, and the first decision is $A(1) = \cA_1(O_1)$. Usually, most algorithms starts by selecting $A(1)=1,\dots,A(K)=K$ (or a permutation of the $K$ arms) in the $K$ first steps (\eg, Algorithm~\ref{algo:2:indexPolicy}).
% as illustrated for example in Algorithm~\ref{algo:2:indexPolicy} below.
%
An algorithm is said to be \emph{deterministic} if it does not depend on the external randomness $U_0,U_1,\dots$, but in this thesis \textbf{we only use non-deterministic algorithms}
(in particular, index policies need $U_t$ to break ties in the $\argmax$, see Algorithm~\ref{algo:2:indexPolicy} below).

@@

\begin{definition}[Kullback-Leibler divergence]\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The KL divergence between two distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
% \vspace*{-20pt}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

Following this Definition~\ref{def:2:KLDivergence} for two distributions
$d_1=\Pr(x|\eta_1)$ and $d_2=\Pr(x|\eta_2)$,
we can write $\KL( \eta_1, \eta_2 ) = \KL(\Pr(x|\eta_1), \Pr(x|\eta_2))$, which can be simplified to use only the two parameters $\eta_1$ and $\eta_2$ of $d_1$ and $d_2$,
and $\mu_1 = \E_{\eta_1}[X]$ the mean of distribution $d_1$:
%
\begin{equation}\label{eq:2:Kullback-LeiblerDivergenceExpFamily2}
    \KL\left( \eta_1, \eta_2 \right) = \E_{d_1}[ (\eta_1 - \eta_2) X ] - A(\eta_1) + A(\eta_2) = (\eta_1 - \eta_2) \mu_1 - A(\eta_1) + A(\eta_2),
\end{equation}

Without diving more in the details of exponential families,
it is interesting to illustrate this definition and the notations with two important examples:

@@

\begin{itemize}
    \item
On the one hand, from a numerical simulation point-of-view, when we run a finite number of repetitions of the same stochastic experiment, if we want to compute and visualize the regret, we can either use the definition with the rewards, or the decomposition and simply sum the gaps $\Delta_k$ with the number of sub-optimal draws.
Both quantities are indeed equal \emph{in expectation}, but with only a finite number of trajectories and observations, the first estimate is more noisy than the second one, since the randomness on the rewards is (partially) removed in the decomposition \eqref{eq:2:RegretDecomposition} (thanks to the conditioning on $O_t$ done in the proof).
In our library SMPyBandits, we implement both estimators, and all values of regret used in this thesis are based on the one using the decomposition of Lemma~\ref{lem:2:RegretDecomposition}, because it gives faster convergence, and also ``smoother'' plots.
\label{remark:2:moreAccurateCountofRegretForSimulations}

    \item
On the other hand, this decomposition is necessary as theoretical analyses of the regret of (single-player) MAB algorithms are usually based on controlling the suboptimal draws $N_k(T)$ and not directly the regret $R_T^{\cA}$.
Even if we prove that controlling $N_k(T)$ is no longer sufficient to obtain low regret for the multi-player case, we extend this decomposition in Chapter~\ref{chapter:3} (in Lemma~\ref{lem:5:DecompositionRegret}), and it is the first quantity we prove to be bounded by $\bigO{\log(T)}$ when we prove the regret upper-bound of our proposal \MCTopM-\klUCB.
\end{itemize}

@@

\begin{theorem}\label{thm:2:firstLogTLowerBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    No algorithm $\cA$ can achieve a (mean) regret $R_T^{\cA}(I) = \smallO{c_I \log(T)}$ for all problem instances $I \in \cI$,
    where this notation means that the ``constant'' $c_I$ can depend on the problem instance $I$ (\eg, on the means and on $K$) but not on the time horizon $T$.
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

This family is non-empty as it contains for instance the $\UCB_1$ algorithm, as its regret is proven to be logarithmic on any problem instance with bounded rewards (in Theorem~\ref{thm:2:UCBregretBound} below).
Now we can state the second logarithmic lower-bound, for algorithms in this family:

@@

\begin{theorem}\label{thm:2:forSecondLogTLowerBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    Let $\cA$ denote a uniformly efficient algorithm,
    and $I$ an arbitrary problem instance.
    \begin{itemize}
        \item
        The bound from Theorem~\ref{thm:2:secondLogTLowerBound} holds with,
        % \begin{equation}\label{eq:2:forSecondLogTLowerBound}
        $C_I = \mu^* (1 - \mu^*) \sum\limits_{k: \Delta_k > 0} \frac{1}{\Delta_k}$.
        % \end{equation}
        \item
        For any $\varepsilon>0$, the bound from Theorem~\ref{thm:2:secondLogTLowerBound} also holds with,
        % \begin{equation}\label{eq:2:forSecondLogTLowerBound2}
        $C_I = \sum\limits_{k: \Delta_k > 0} \frac{\Delta_k}{\KL(\mu_k, \mu^*)} - \varepsilon$.
        % \end{equation}
    \end{itemize}
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

% ----------------------------------------------------------------------------
\textbf{Policies for adversarial bandits.}
%
Another important family of bandits algorithms are the algorithms proposed for the adversarial setting \cite{Auer02NonStochastic}.
The first of them is the Exp3 policy, for ``\emph{Exp}onential weights for \emph{Exp}loitation and \emph{Exp}loiration'',
which is an example of the well-known \emph{multiplicative weights update algorithm}\footnote{~For more details see
% on Wikipedia \href{https://en.wikipedia.org/wiki/Multiplicative_weight_update_method}{\texttt{en.wikipedia.org/wiki/Multiplicative\_weight\_update\_method}},
this blog post by Jeremy Kun, \href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{JeremyKun.com/2017/02/27/the-reasonable-}}\\
\href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{effectiveness-of-the-multiplicative-weights-update-algorithm}},
which comes with an interactive online demonstration to help understanding the behavior of the algorithm, hosted at \href{https://j2kun.github.io/mwua/index.html}{\texttt{j2kun.GitHub.io/mwua/}} (accessed the 24th of May 2019). More details can be found in the survey \cite{Arora2012multiplicative}.},
a family of algorithms developed since the 1950s, and applied to a wide range of different domains.
% https://smpybandits.github.io/docs/Policies.Exp3.html
%
The Exp3 policy maintains weights $w_k(t)$ on the $K$ arms, and at time $t$ it samples an arm from the distribution which is a mixture of $\bm{w}(t)=[w_1(t),\dots,w_K(t)]$ and the uniform distribution:
$\Pr(A(t)=k) = (1-\gamma) w_k(t) + \gamma \frac{1}{K}$.
The weights are initially uniform, \ie, $w_k(0)=\frac{1}{K}$, and then at each time after observing a reward $Y_{k,t}$ from arm $k$, Exp3 updates the weight of the chosen arm multiplicatively, using $w_k(t+1) = w_k(t) \times \exp(\widehat{r_k}(t) / (\gamma N_k(t)))$. The weights are then renormalized, the have a sum $\sum_k w_k(t+1) = 1$.
This notation $\widehat{r_k}(t)$ means an unbiased estimator of the reward $r(t)=Y_{k,t}$, that is, $\widehat{r_k}(t) = \Pr(A(t)=k)$.
%
Like for the $\varepsilon$-greedy policy, the parameter $\gamma$ can be constant, or can be a decreasing sequence $(\gamma_t)_t$.
%
It was proved in \cite{Auer02NonStochastic} that Exp3 with constant temperature parameter $\gamma$ achieves $R_T = \bigO{\sqrt{K T \log(T)}}$ problem-independent regret,
while using a decreasing sequence, such as $\gamma_t = \sqrt{\log(K) / K} \log(t) / t$ gives an order-optimal problem-independent regret upper-bound of $\bigO{\sqrt{K T}}$.
A short proof of this results and more details on Exp3 are given in \cite{Bubeck12}.
%
Many other policies have been proposed for adversarial bandits, like Follow-the-Perturbed-Leader or Follow-the-Regularized-Leader for linear adversarial bandits, but we do not study them in this thesis, and we refer to Parts~III and VI of \cite{LattimoreBanditAlgorithmsBook} for more details.

@@

\textbf{Illustrative example:}
%
For instance, if you know that the considered problem can either have $K$ arms with very close means, or one optimal arm far away from the other, two versions of a $\UCB_1$ algorithms will perform quite differently in the two problems:
using a large $\alpha$, \ie, favoring exploration, will give low regret in the first case,
while using a low $\alpha$, \ie, favoring exploitation, will give low regret in the second case.
One approach can be to use an intermediate value, as $\alpha=1/2$ suggested by theory, but another approach could be to consider an aggregated vote of different versions of $\UCB_1$, each running with a different value of $\alpha$ (\eg, in a logarithmic grid), and let another learning algorithm decide which value of $\alpha$ is the best \emph{for the problem at hand}.

>>>>

Multi-Armed Bandits (MAB) models were introduced by Thompson as early as in 1933 \cite{Thompson33}, and later studied from the 1950s by Robbins \cite{Robbins52} and others.
This family of models was first proposed for clinical trials, and later applied to a wide range of different problems.
Their name refers to one-armed bandits found in casinos, as shown in Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
% The seminal paper by Lai and Robbins \cite{LaiRobbins85} was the first to introduce the measure of performance widely used in the literature, the notion of \emph{regret}, and to analyze a lower-bound on the performance of any algorithm against a fixed problem.
% Since then, many efficient algorithms were proposed, XXX
%
% In all this thesis, we only consider one-dimensional exponential families, and in particular Bernoulli distributions.
We formally define the model and the main notations used in all this document in Section~\ref{sec:2:notations}.
%
Multi-armed bandits have been applied to various fields, from recommender systems and information retrieval to healthcare and finance. We discuss some applications in Section~\ref{sec:2:applicationsofStochasticMAB}.
In the rest of this thesis, we focus on cognitive radio, an important field of research where bandits has already been applied with success in the last ten years.

@@

A rich research literature has produced many different algorithms tackling the MAB problem, as suggested by the length of the reference book \cite{LattimoreBanditAlgorithmsBook}.
We give in Section~\ref{sec:2:famousMABalgorithms} an overview of the main families of algorithms designed to tackle stochastic MAB problems, and we mainly focus on the $\UCB_1$, \klUCB{} and Thompson sampling algorithms, as they are used in the rest of the manuscript.
% We present SMPyBandits in the next Chapter~\ref{chapter:3}, a Python open-source library, designed to run numerical simulations of MAB problems, one of the main contributions of this thesis.
% We use it to compare the empirical performances of the most famous and most efficient algorithms, in terms of regret and by evaluating them on different problems.
% %
% In the point-of-view of numerical simulations or real-world applications of bandits on embedded systems, one must also take care of two other measures of performance: time and memory complexities, which can vary greatly between different families of algorithms.
% We also empirically compare different algorithms for these two measures in Section~\ref{sec:3:timeAndMemoryCosts}.
%
% We then give regret upper-bounds of the two main algorithms used in this thesis, $\UCB_1$ and \klUCB, to highlight the difference between their finite-time regret upper-bounds. They are order-optimal as they attain a logarithmic regret.
% proving their order optimality.
% (\ie, their regret upper-bound asymptotically matches the lower-bound up-to a constant factor).
% Moreover, \klUCB{} is known to be optimal for a wide range of arms distributions.
% A sketch of the proof of both results is given in Appendix.
%
Finally, we explore in Section~\ref{sec:2:chooseYourPreferredBanditAlgorithm} the problem of algorithm selection: as there are many different algorithms, all having different qualities and properties, how one can decide which algorithm to use for a particular application?
We present one of the first contributions made during my PhD \cite{Besson2018WCNC}, consisting of an efficient algorithm for aggregating algorithms,
and we illustrate its performance empirically.
% , but no satisfactory theoretical result is given, as the bound we had hoped for has been proved unachievable since then.

@@

A MAB is a model of a decision-making game, where a \emph{player} has to sequentially select one action in a (usually finite) set of actions, and only receives a (random) feedback about the selected action, also called a \emph{reward}.
\textbf{We always consider $K\in\N, K\geq2$ arms, and discrete times $t\in\N^*$}.
When the player decides at time $t$ to play (or pull) the arm $A(t)\in\{1,\dots,K\}=[K]$, she receives a reward $r(t) = Y_{A(t),t}\in\R$, and so on for a finite number of steps $t$ (or rounds), from $t=1$ until $t=T$ for an \emph{horizon} $T$.
%
A commonly studied goal for the player is to maximize its sum of received rewards, $\sum_{t=1}^T r(t)$
(or its \emph{mean} in stochastic and stationary models).

@@

The most common objective for the player is to maximize its expected rewards $\E[\sum_{t=1}^T r(t)] = \sum_{t=1}^T \E[Y_{A(t),t}]$.
The expectancy $\E[\bullet]$ is taken on the rewards $(Y_{k,t})_{k,t}$, as well as on the external random variables $(U_s)_s$ (\ie, on the algorithm's decisions).
%
Other commonly studied objectives include best-arm identification (BAI) \cite{audibert2010best}.

@@

The webpage looks like Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem} below.
The arms follow Bernoulli distributions, \ie, $\nu_k = \mathcal{B}(\mu_k)$, of unknown means $\mu_k\in[0,1]$, and your goal in this interactive demonstration is to obtain the highest possible cumulated reward in $T=100$ steps, \ie, to maximize $\sum_{t=1}^{100} r(t)$.
Your decisions are made sequentially: at time $t$, you pick one of the arms, $A(t) \in\{1,2,3,4,5\}$, then the demo shows the random reward obtained from this (virtual) casino machine (in \textcolor{gold}{yellow}), \ie, the binary value $r(t)\in\{0,1\}$, that is sampled \iid{} from $\nu_k$.
%
The UI of the demo also shows the current value of $t$ (``total plays'') and $\sum_{s=1}^t r(s)$ the ``total reward''.
For each arm, we show the sum of rewards obtained from that arm, \ie, $X_k(t) = \sum_{s=1}^t r(s) \mathbbm{1}(A(s) = k)$, in the ``Rewards'' line, and the number of pulls of that arm, \ie, $N_k(t) = \sum_{s=1}^t \mathbbm{1}(A(s) = k)$ in the ``Pulls'' line.
%
The demo also shows the estimated probability of each arm, that is $\widehat{\mu_k}(t) = X_k(t) / N_k(t)$ (when $N_k(t)>0$), in the ``Estimated Probs'' line.

In the first Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem}, the current state of the game is shown at time $t=24$.
% Currently at time $t=24$ (out of $T=100$ total time steps),
At this step, the player has collected a sum of rewards of $14$, by observing $X(t) = [6,2,2,2,2]$ rewards of value $1$ in the $K=5$ different arms. Arms were sampled $N(t) = [8,4,4,4,4]$ times, meaning that the value $0$ was seen respectively $[2,2,2,2,2]$ times, and currently arm $1$ appears to be the best one. The true means of the arms are $\bm{\mu}=[0.6, 0.2, 0.55, 0.7, 0.5]$, and (much) more samples are needed before the player can accurately identify arm $4$ as the best arm.
%
In the second Figure~\ref{fig:2:example_of_a_5_arm_bandit_problem__step100} below, we display the result of an example of run, when the player was following the $\UCB_1$ algorithm from \cite{Auer02} (we present it below in Section~\ref{sub:2:IndexPolicies}).
After $T=100$ steps, the player obtained a cumulated reward of $56$, by playing mostly arms $4,3,5,1,0$ (in decreasing order of number of plays). The empirical means $\widehat{\mu_k}(T)$ correctly identify the best arm (arm $4$), but do not correctly rank the arms as arms $1$ and $3$ obtained means of $\widehat{\mu_1}(T) = 0.5 < \widehat{\mu_3}(T)=0.6$ but the true means are $\mu_3 = 0.55 < \mu_1 = 0.6$.
Other examples of such results, for different algorithms, and $T=100$ or $T=10000$, are given in Section~\ref{sub:2:shortNumericalExperiments}.

@@

A \emph{bandit algorithm} $\cA$ is also referred to as a strategy or a policy.
% and sometimes it is denoted by $\pi$ or $\rho$ in the literature.
The algorithm $\cA$ selects an arm $A(t)$ at time $t$, possibly by using the past observations and the past external randomness.
As shown below, being oblivious to the past yields very poor performance (cf. the pure exploration policy in Section~\ref{sub:2:naiveSimpleStrategies}), so efficient policies indeed depend on the successive feedbacks.
%
More formally, an algorithm can be defined as a sequence of \emph{measurable} functions $(\cA_t)_{t\geq1}$,
where $\cA_t$ maps the past observations $O_t = (U_0, Y_{A(1),1}, U_1, \dots, Y_{A(t-1),t-1}, U_{t-1})$
to an arm $\cA_t(O_t) = A(t) \in[K]$
(we remind that we denote $r(s) = Y_{A(s),s}$ the $s$-th reward).
The initial information is reduced to $O_1 = (U_0)$, and the first decision is $A(1) = \cA_1(O_1)$. Usually, most algorithms starts by selecting $A(1)=1,\dots,A(K)=K$ (or a permutation of the $K$ arms) in the $K$ first steps (\eg, Algorithm~\ref{algo:2:indexPolicy}).
% as illustrated for example in Algorithm~\ref{algo:2:indexPolicy} below.
%
An algorithm is said to be \emph{deterministic} if it does not depend on the external randomness $U_0,U_1,\dots$, but in this thesis \textbf{we only use non-deterministic algorithms}
(in particular, index policies need $U_t$ to break ties in the $\argmax$, see Algorithm~\ref{algo:2:indexPolicy} below).

@@

\begin{definition}\label{def:2:KLDivergence}
\begin{leftbar}[defnbar]  % XXX leftbar defnbar, comment if needed
    The \emph{Kullback-Leibler divergence} between distributions $d_1$ and $d_2$ is defined as
    \begin{align}\label{eq:2:Kullback-LeiblerDivergenceExpFamily1}
        \KL\left(d_1, d_2\right) = \int d_1 \log\left(\frac{d_2}{d_1}\right) \lambda(\mathrm{d}x) = \E_{d_1} \left[ \log\left(\frac{d_2}{d_1}\right) \right] \in \mathbb{R}\cup\{\pm\infty\}.
    \end{align}
% \vspace*{-5cm}  % FIXME vspace negative to compensate the \end{leftbar}
\end{leftbar}  % XXX leftbar defnbar, comment if needed
\end{definition}

Following this Definition~\ref{def:2:KLDivergence} for two distributions
$d_1=\Pr(x|\eta_1)$ and $d_2=\Pr(x|\eta_2)$,
we can write $\KL( \eta_1, \eta_2 ) = \KL(\Pr(x|\eta_1), \Pr(x|\eta_2))$, which can be simplified to use only the two parameters $\eta_1$ and $\eta_2$ of $d_1$ and $d_2$,
and $\mu_1 = \E_{\eta_1}[X]$ the mean of distribution $d_1$,
%
% \begin{equation}\label{eq:2:Kullback-LeiblerDivergenceExpFamily2}
$\KL\left( \eta_1, \eta_2 \right) = \E_{d_1}[ (\eta_1 - \eta_2) X ] - A(\eta_1) + A(\eta_2) = (\eta_1 - \eta_2) \mu_1 - A(\eta_1) + A(\eta_2)$.
% \end{equation}
%
Without diving more in the details of exponential families,
it is interesting to illustrate this definition and the notations with two important examples:

@@

% \begin{itemize}
%     \item
On the one hand, from a numerical simulation point-of-view, when we run a finite number of repetitions of the same stochastic experiment, if we want to compute and visualize the regret, we can either use the definition with the rewards, or the decomposition and simply sum the gaps $\Delta_k$ with the number of sub-optimal draws.
Both quantities are indeed equal \emph{in expectation}, but with only a finite number of trajectories and observations, the first estimate is more noisy than the second one, since the randomness on the rewards is (partially) removed in the decomposition \eqref{eq:2:RegretDecomposition} (thanks to the conditioning on $O_t$ done in the proof).
In our library SMPyBandits, we implement both estimators, and all values of regret used in this thesis are based on the one using the decomposition of Lemma~\ref{lem:2:RegretDecomposition}, because it gives faster convergence, and also ``smoother'' plots.
\label{remark:2:moreAccurateCountofRegretForSimulations}

    % \item
On the other hand, this decomposition is necessary as theoretical analyses of the regret of (single-player) MAB algorithms are usually based on controlling the suboptimal draws $N_k(T)$ and not directly the regret $R_T^{\cA}$.
Even if we prove that controlling $N_k(T)$ is no longer sufficient to obtain low regret for the multi-player case, we extend this decomposition in Chapter~\ref{chapter:3} (in Lemma~\ref{lem:5:DecompositionRegret}), and it is the first quantity we prove to be bounded by $\bigO{\log(T)}$ when we prove the regret upper-bound of our proposal \MCTopM-\klUCB.
% \end{itemize}

@@

\begin{theorem}\label{thm:2:firstLogTLowerBound}
\begin{leftbar}[theorembar]  % XXX leftbar theorembar, comment if needed
    \emph{No algorithm $\cA$ can achieve} a (mean) regret $R_T^{\cA}(I) = \smallO{c_I \log(T)}$ for all problem instances $I \in \cI$,
    where this notation means that the ``constant'' $c_I$ can depend on the problem instance $I$ (\eg, on the means and on $K$) but not on the time horizon $T$.
\end{leftbar}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

This family is non-empty as it contains for instance the $\UCB_1$ algorithm, since its regret is proven to be logarithmic on any problem instance with bounded rewards (in Theorem~\ref{thm:2:UCBregretBound} below).
Now we can state the second logarithmic lower-bound, for algorithms in this family:

@@

\begin{theorem}\label{thm:2:forSecondLogTLowerBound}
\begin{leftbarnospace}[theorembar]  % XXX leftbar theorembar, comment if needed
    Let $\cA$ denote a uniformly efficient algorithm,
    and $I$ an arbitrary problem instance.
    \begin{itemize}
        \item
        The bound from Theorem~\ref{thm:2:secondLogTLowerBound} holds with,
        % \begin{equation}\label{eq:2:forSecondLogTLowerBound}
        $C_I = \mu^* (1 - \mu^*) \sum\limits_{k: \Delta_k > 0} \frac{1}{\Delta_k}$.
        % \end{equation}
        \item
        For any $\varepsilon>0$, the bound from Theorem~\ref{thm:2:secondLogTLowerBound} also holds with,
        % \begin{equation}\label{eq:2:forSecondLogTLowerBound2}
        $C_I = \sum\limits_{k: \Delta_k > 0} \frac{\Delta_k}{\KL(\mu_k, \mu^*)} - \varepsilon$.
        % \end{equation}
    \end{itemize}
\end{leftbarnospace}  % XXX leftbar theorembar, comment if needed
\end{theorem}

@@

% ----------------------------------------------------------------------------
\textbf{Policies for adversarial bandits.}
%
Another important family of bandits algorithms are the algorithms proposed for the adversarial setting \cite{Auer02NonStochastic}.
The first of them is the Exp3 policy, for ``\emph{Exp}onential weights for \emph{Exp}loitation and \emph{Exp}loiration'',
which is an example of the well-known \emph{multiplicative weights update algorithm}\footnote{~For more details see
% on Wikipedia \href{https://en.wikipedia.org/wiki/Multiplicative_weight_update_method}{\texttt{en.wikipedia.org/wiki/Multiplicative\_weight\_update\_method}},
this blog post by Jeremy Kun, \href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{JeremyKun.com/2017/02/27/the-reasonable-}}\\
\href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{effectiveness-of-the-multiplicative-weights-update-algorithm}},
which comes with an interactive online demonstration to help understanding the behavior of the algorithm, hosted at \href{https://j2kun.github.io/mwua/index.html}{\texttt{j2kun.GitHub.io/mwua/}} (accessed the 24th of May 2019). More details can be found in the survey \cite{Arora2012multiplicative}.},
a family of algorithms developed since the 1950s, and applied to a wide range of different domains.
% https://smpybandits.github.io/docs/Policies.Exp3.html
%
The Exp3 policy maintains weights $w_k(t)$ on the $K$ arms, and at time $t$ it samples an arm from the distribution which is a mixture of $\bm{w}(t)=[w_1(t),\dots,w_K(t)]$ and the uniform distribution:
$\Pr(A(t)=k) = (1-\gamma) w_k(t) + \gamma \frac{1}{K}$.
The weights are initially uniform, \ie, $w_k(0)=\frac{1}{K}$, and then at each time after observing a reward $Y_{k,t}$ from arm $k$, Exp3 updates the weight of the chosen arm multiplicatively, using $w_k(t+1) = w_k(t) \times \exp(\widehat{r_k}(t) / (\gamma N_k(t)))$. The weights are then renormalized, to have a sum $\sum_k w_k(t+1) = 1$.
This notation $\widehat{r_k}(t)$ means an unbiased estimator of the reward $r(t)=Y_{k,t}$, that is, $\widehat{r_k}(t) = \Pr(A(t)=k)$.
%
Like for the $\varepsilon$-greedy policy, the parameter $\gamma$ can be constant, or can be a non-increasing sequence $(\gamma_t)_t$.
%
It was proved in \cite{Auer02NonStochastic} that Exp3 with constant parameter $\gamma$ achieves $R_T = \bigO{\sqrt{K T \log(T)}}$ problem-independent regret,
while using a decreasing sequence, such as $\gamma_t = \sqrt{\log(K) / K} \log(t) / t$ gives an order-optimal problem-independent regret upper-bound of $\bigO{\sqrt{K T}}$.
A short proof of this results and more details on Exp3 are given in \cite{Bubeck12}.
%
Many other policies have been proposed for adversarial bandits, like Follow-the-Perturbed-Leader or Follow-the-Regularized-Leader for linear adversarial bandits,
% but we do not study them in this thesis,
and we refer to Parts~III and VI of \cite{LattimoreBanditAlgorithmsBook} for more details.

@@

\textbf{Illustrative example:}
%
For instance, if you know that the considered problem can either have $K$ arms with very close means, or one optimal arm far away from the other, two versions of $\UCB_1$ will perform quite differently in the two problems:
using a large $\alpha$, \ie, favoring exploration, will give low regret in the first case,
while using a low $\alpha$, \ie, favoring exploitation, will give low regret in the second case.
One approach can be to use an intermediate value, as $\alpha=1/2$ suggested by theory, but another approach could be to consider an aggregated vote of different versions of $\UCB_1$, each running with a different value of $\alpha$ (\eg, in a logarithmic grid), and let another learning algorithm decide which value of $\alpha$ is the best \emph{for the problem at hand}.
